EvoSwarm: A Cloud-Native Architecture for the Reproducible Execution of Multi-population Optimization Algorithms

### 2020-08-21 draft.tex
Fixes issue #17

However, in the field of evolutionary algorithms, this has been a slow
realization, from the origins where Salza and Ferrucci
\cite{salza2016approach,salza2016develop} proposed an architecture
that carried evolutionary algorithms to the cloud, and which were
developed further in their next paper \cite{de2017parallel}. These
papers started to introduce cloud native aspects such as the use of
messaging queues and CoreOS as an operating system designed from the
ground up for container utilization in the area of evolutionary
algorithms. The way these containers are managed, however, is rather
classical in nature from the point of view of distributed EC: a
master-worker approach, communicated using RabbitMQ (a messaging
queue), where replicated workers perform the tasks in parallel. They
called their approach AMQPGA, inspired by the protocol, AMQP, used by
RabbitMQ, the messaging system. This classical approach has been also
followed by other authors by using other EA ``classical'' topologies
in the cloud: Dziurzanski et al. \cite{dziurzanski2020scalable} adapts an island model to
a container architecture, in a system that is functionally equivalent
to classical models, but with the added advantage of being easy to
deploy and run.

@@

EvoSwarm describes the deployment of loosely coupled processes in a consistent
and portable way. The term cloud-native describes these container-based
environments because we can deploy them on an infrastructure that abstracts the
underlying compute, storage, and networking primitives. This infrastructure does
not need to be a cloud provider; it could be a developer's laptop, a single
server, or a production environment. In this section, we give specific details 
about how we deployed  EvoSwarm in a cloud-setting for running the experiments 
within this paper.

@@

  \item[A Kubernetes cluster] provides auto-scaling and management of containerized
  applications, but it needs additional configuration for provisioning. All major
  cloud-providers have Kubernetes services.

@@

@@

\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{aws_ec2}
  \caption{AWS EC2 configuration for the deployment of EvoSwarm containers.
   }
  \label{fig:aws-configuration}
\end{figure}
%

@@

To measure the speedup obtained by the addition of workers, we will use the
number of evaluations per second, instead of the time required to finish an
instance of a problem. The reason for this is that in some cases reaching the
desired target can require less work, because we are using a stochastic search. 
We are going to measure the total increase of work done in a second by adding more workers.
The baseline for the comparison will be the algorithm with a single worker, 
and the minimum amount of work available will be five populations. 
This number is selected because the controller needs at least
three populations to migrate individuals between them, and we add two more to
keep even a single worker busy all the time. Figure \ref{fig:spworker:evalspersecond} 
shows the number of function evaluations per second, achieved by a certain number of
workers, for ten and twenty dimensions of $f_4$.

@@

  \begin{tabular}{c|r|r|r|r|}
  \hline
  \multicolumn{1}{|l|}{Dimension} & \multicolumn{1}{c|}{2W} & \multicolumn{1}{c|}{4W} & \multicolumn{1}{c|}{8W} & \multicolumn{1}{c|}{16W}  \\ \hline
  \multicolumn{5}{| c|}{5 Populations} \\ \hline
  \multicolumn{1}{|c|}{10}        & \textbf{2.15} & 3.61 & 3.75 & 3.96     \\ \hline
  \multicolumn{1}{|c|}{20}        & \textbf{2.09} & 3.53 & 3.83 & 3.95     \\ \hline
  \multicolumn{5}{| c|}{10 Populations} \\ \hline
  \multicolumn{1}{|c|}{10}        & 1.91           & 3.82 & 7.67 & 8.25     \\ \hline
  \multicolumn{1}{|c|}{20}        & 1.97           & 3.97 & 7.90 & 8.90     \\ \hline
  \multicolumn{5}{| c|}{20 Populations} \\ \hline
  \multicolumn{1}{|c|}{10}        & 1.96           & \textbf{4.05} & \textbf{8.29} & \textbf{14.50}     \\ \hline
  \multicolumn{1}{|c|}{20}        & 2.00           & \textbf{4.09} & \textbf{8.18} & \textbf{15.16}     \\ \hline
  \end{tabular}
\end{table}
%
As expected, within the same number of populations, there is a linear speedup
when we add more workers. The speedup is marginal in those cases where there 
are fewer populations than workers.  The reason for this is that at any given time, 
there are at least two populations in the message queue, leaving some of the workers 
waiting for populations to be available in the queue.

@@

Table~\ref{tab:speedup:test} shows the speedup obtained taking as the baseline
the previous configuration, for example, when increasing from four to eight
workers, and the p-value for the left-tailed Wilcoxon test. There was an
increment in all cases. Again, the best effect is achieved when the number of
populations is higher than the number of workers. Configurations following this
basic rule doubled the speed or where near it. Using five populations, we
obtained the best and worst speedup depending on the number of workers (shown in
underline and bold, respectively).

@@

%
\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{secs_per_instance}
  \caption{Boxplot of the median time required for the execution of 30 
  instances of function $f_4$, in seconds, different number of workers and populations sizes. The y-axis is shown in log scale.  }
  \label{fig:spworker:time}
\end{figure}
%
\begin{table}[tbp]
  \caption{The median of the time required to finish an instance of the function $f_4$, 30 runs
  with different number of workers and populations}
  \label{tab:time}
  \vspace{0.25cm}
    \centering
    \begin{tabular}{ccrrr}
    \hline
              &         & 5 Populations    & 10 Populations & 20 Populations \\
     Dimension& Workers & Median of Time   & Median of Time & Median of Time  \\
    \hline
          10  & 1w      & 22.24  & 20.48           & 24.55  \\
              & 2w      & 10.31  & \textbf{10.00} (2.05)    & 12.79 \\
              & 4w      &  6.23  & \textbf{4.75} (2.10)     &  5.70 \\
              & 8w      &  5.86  &  \textbf{2.33} (2.03)           & 2.74 \\
              & 16w     &  5.96*  &  2.33*           & {\ul 1.59} \\

@@

@@

In Table~\ref{tab:time} we can see that as expected, and in all cases, 
increasing the number of workers reduces the time required for the execution of an instance, 
mainly when using the same parameters (number of populations and number of workers), that is, performance
scales with the number of workers for a fixed problem and number of
sub-populations used.

However, when we compare the two options, we find some subtle differences: When
using one or two workers, using 5 populations takes less time in most cases. In
Table~\ref{tab:time} results with p-value of the two-tailed Wilcoxon
signed-ranks test was grater than the level of significance of 0.05 (i.e.,
accepting the null hypothesis of inequality) are marked in boldface. A reason
for this could be that when having more populations, the migration of
individuals takes more time from the bdyeginning to reach to all of them. For
instance, when a single worker has 10 populations, only after processing every
one of them, it will start receiving mixed populations. Although, in all cases,
and even with a single worker, populations are heterogeneous, individuals
resulting from the mixing of populations take more time to propagate. With four
workers, almost the same results can be achieved regardless of our selection
between five or ten populations. Also, there is not much difference between 4
and 8 workers, when using five populations, which might indicate that some
workers are idling while populations are being processed by other workers. In
general, we can observe that keeping a ratio between the number of workers and
populations a bit over 1:1 exploits better the computing resources available.
This also decreases the relation between the overhead needed to migrate and
process incoming and outgoing population and the total computing time: more
populations will need more time to do that.

%
%

@@

% In this paper, we have presented the EvoSwarm framework that implements
% an event-based, cloud-native architecture suitable for hosting 
% multi-population optimization algorithms. The framework is
% based in industry-standard development tools such as Docker and Docker
% Compose, making it easy to develop and deploy multi-algorithm distributed
% experiments.

% The set up of this kind of algorithms will have a big influence on the
% results, since diversity is important for this kind of algorithms and
% its creation and destruction play a big part of the algorithm's
% performance. One of the main levers to control it in EvoSwarm is via
% the number of workers that will be allocated from the beginning, which
% is why we have performed an experiment mainly intended as a proof of
% concept, but also devoted to show how two parameters in the framework,
% number of workers and number of populations used to initially seed the
% system, interact with each other. The conclusion of this experiment is
% that there is an interesting interplay, with the most benefit being
% obtained with a number of populations equal to the number of workers
% or slightly inferior or superior, which brings algorithmic benefit
% (experiment taking less time because it needs less evaluations) and
% also performance benefits (experiment taking less time because
% evaluations are being split between different workers.

% % this is not a conclusion. The conclusion is an answer to the
% % research question. This works, and works precisely for the reasons
% % we say it works. - JJ
% % Again, revise search questions and explain answers - JJ
% As a conclusion, EvoSwarm is a framework designed and tested according
% to industry standards which is able to bring many benefits to the
% bioinspired algorithm community, by showing the possibility of
% creating new, high performance, mixed algorithms, which can also be
% easily deployed to the cloud, or locally to systems that simply need
% to have Docker and {\tt docker-compose} installed.

This work also opens new possibilities. From the design point of view, it
would be interesting to try and use automatic scaling, instead of
setting the number of workers by hand. This would need a certain
amount of redesign, the most important of which would be using
container orchestration systems such as Kubernetes or Docker
Swarm. Instead of setting a fixed number of workers, only the maximum
amount would need to be established, which could be done directly or
by setting an evaluation budget.

@@

\section*{References}

>>>>

However, in the field of evolutionary algorithms, this has been a slow
realization, from the origins where Salza and Ferrucci
\cite{salza2016approach,salza2016develop} proposed an architecture that carried
evolutionary algorithms to the cloud, and which were developed further in their
next paper \cite{de2017parallel}. These papers started to introduce cloud native
aspects such as the use of messaging queues and CoreOS as an operating system
designed from the ground up for container utilization in the area of
evolutionary algorithms. The way these containers are managed, however, is
rather classical in nature from the point of view of distributed EC: a
master-worker approach, communicated using RabbitMQ (a messaging queue), where
replicated workers perform the tasks in parallel. They called their approach
AMQPGA, inspired by the protocol, AMQP, used by RabbitMQ, the messaging system.
Other authors have also followed this classical approach by using other EA
``classical'' topologies in the cloud: Dziurzanski et al.
\cite{dziurzanski2020scalable} adapts an island model to a container
architecture, in a system that is functionally equivalent to classical models,
with the advantage of the auto-scaling of islands, and being easy to deploy and
run.

@@

EvoSwarm describes the deployment of loosely coupled processes in a consistent
and portable way. The term cloud-native describes these container-based
environments because we can deploy them on an infrastructure that abstracts the
underlying compute, storage, and networking primitives. This infrastructure does
not need to be a cloud provider; it could be a developer's laptop, a single
server, or a production environment. In this section, we give specific details
about how we deployed EvoSwarm in a cloud-setting for running the experiments
within this paper. Running the cloud-native application in the cloud has the
benefit of a multi-instance deployment, where we can run containers in several
instances executing at the same time, not only in a single machine.

@@

  \item[A Kubernetes cluster] provides auto-scaling and management of
  containerized applications, but it needs additional configuration for
  provisioning. All major cloud-providers have Kubernetes services. This option
  was used by S. Zhao et al. \cite{zhao2019cloud} to run a multi-island
  algorithm, and Dziurzanski et al. \cite{dziurzanski2020scalable} later refined
  it.

@@

\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{aws_ec2}
  \caption{AWS EC2 configuration for the deployment of EvoSwarm containers.
   }
  \label{fig:aws-configuration}
\end{figure}

@@

@@

When comparing the results in this section, we performed the Wilcoxon Signed-Rank test, with a
significance level of 5\%. We indicate with an asterisk those results that are
not significant.

To measure the speedup obtained by the addition of workers, we will use the
number of evaluations per second, instead of the time required to finish an
instance of a problem. The reason for this is that in some cases reaching the
desired target can require less work, because we are using a stochastic search. 
We are going to measure the total increase of work done in a second by adding more workers.

@@

  \begin{tabular}{c|r|r|r|r|}
  \hline
  \multicolumn{1}{|l|}{Dimension} & \multicolumn{1}{c|}{2W} & \multicolumn{1}{c|}{4W} & \multicolumn{1}{c|}{8W} & \multicolumn{1}{c|}{16W}  \\ \hline
  \multicolumn{5}{| c|}{5 Populations} \\ \hline
  \multicolumn{1}{|c|}{10}        & \textbf{2.15} & 3.61 & 3.75 & 3.96     \\ \hline
  \multicolumn{1}{|c|}{20}        & \textbf{2.09} & 3.53 & 3.83 & 3.95     \\ \hline
  \multicolumn{5}{| c|}{10 Populations} \\ \hline
  \multicolumn{1}{|c|}{10}        & 1.91           & 3.82 & 7.67 & 8.25     \\ \hline
  \multicolumn{1}{|c|}{20}        & 1.97           & 3.97 & 7.90 & 8.90     \\ \hline
  \multicolumn{5}{| c|}{20 Populations} \\ \hline
  \multicolumn{1}{|c|}{10}        & 1.96           & \textbf{4.05} & \textbf{8.29} & \textbf{14.50}     \\ \hline
  \multicolumn{1}{|c|}{20}        & 2.00           & \textbf{4.09} & \textbf{8.18} & \textbf{15.16}     \\ \hline
  \end{tabular}
\end{table}

The baseline for the comparison will be the algorithm with a single worker, 
and the minimum amount of work available will be five populations. 
This number is selected because the controller needs at least
three populations to migrate individuals between them, and we add two more to
keep even a single worker busy all the time. Figure \ref{fig:spworker:evalspersecond} 
shows the number of function evaluations per second, achieved by a certain number of
workers, for ten and twenty dimensions of $f_4$.

%
As expected, within the same number of populations, there is a linear speedup
when we add more workers. The speedup is marginal in those cases where there 
are fewer populations than workers.  The reason for this is that at any given time, 
there are at least two populations in the message queue, leaving some of the workers 
waiting for populations to be available in the queue.

@@

Table~\ref{tab:speedup:test} shows the speedup obtained taking as the baseline
the previous configuration, for example, when increasing from four to eight
workers, and the p-value for the Wilcoxon Signed-Rank test. There was an
increment in all cases. Again, the best effect is achieved when the number of
populations is higher than the number of workers. Configurations following this
basic rule doubled the speed or where near it. Using five populations, we
obtained the best and worst speedup depending on the number of workers (shown in
underline and bold, respectively).

@@

@@

%
\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{secs_per_instance}
  \caption{Boxplot of the median time required for the execution of 30 
  instances of function $f_4$, in seconds, different number of workers and 
  populations sizes. The y-axis is shown in log scale.}
  \label{fig:spworker:time}
\end{figure}
%

@@

\begin{table}[tbp]
  \caption{The median of the time required to finish an instance of the function $f_4$, 30 runs
  with different number of workers and populations}
  \label{tab:time}
  \vspace{0.25cm}
    \centering
    \begin{tabular}{ccrrr}
    \hline
              &         & 5 Populations    & 10 Populations & 20 Populations \\
     Dimension& Workers & Median of Time   & Median of Time & Median of Time  \\
    \hline
          10  & 1w      & 22.24  & 20.48           & 24.55  \\
              & 2w      & 10.31  & \textbf{10.00} (2.05)    & 12.79 \\
              & 4w      &  6.23  & \textbf{4.75} (2.10)     &  5.70 \\
              & 8w      &  5.86  &  \textbf{2.33} (2.03)           & 2.74 \\
              & 16w     &  5.96*  &  2.33*           & {\ul 1.59} \\

In Table~\ref{tab:time} we can see that in some cases we have a supralinear
speedup, we highlighted these in boldface and indicated the speedup value in
parenthesis. The best overall times are underlined. As expected, increasing the
number of workers for a fixed dimension and populations reduced the time
required to find a solution, as long as the population number was higher than
the number of workers. The values that had no statistical difference (in the
Wilcoxon test) when increasing the number of workers have an asterisk.

@@

This work also opens new possibilities. From the design point of view, it would
be interesting to use automatic scaling, instead of setting the number of
workers by hand. This would need a certain amount of configuration, as we would
be using container orchestration with auto-scaling capabilities such as
Kubernetes or Docker Swarm. The current orchestration (Amazon ECS) also has
auto-scaling capabilities, but additional configuration is needed; also, we need
to establish the auto-scaling of both containers and Amazons EC2 instances.
Instead of setting a fixed number of workers, only the maximum amount would need
to be established, which could be done directly or by setting an evaluation
budget.

@@

%\section*{References}
